---
title: "Land Lab Garter Snake Analysis in R"
author: "Biana Qiu"
date: "2026-04-27"
format: 
  html:
    toc: true
    toc-depth: 3
    css: BQ_custom.css
    self-contained: true
    code-fold: true
    code-summary: "Open Code"
    code-tools: true
execute: 
  echo: true
  warning: false
  message: false
number-sections: true
---

This is the code ran for Biana Qiu's 2026 senior research project: The effects of habitat variables on site occupancy of the Common Gartersnake (Thamnophis sirtalis) on a restored prairie/wetland complex. Analysis ran on data gathered by Geoffrey R. Smith, Wesley O. Smith and many others throughout 2016-2025 in the Granville Schools Land Lab. Analysis contains single- and multi-season/dynamic occupancy using package 'unmarked.'

### Libraries

```{r}
#install.packages("tidyverse")
library(tidyverse) #for data organization and graphing
#install.packages("magrittr")
#library(magrittr) #for piping and data manipulation
#install.packages("corrplot")
library(corrplot) #for correlation plots
#install.packages("unmarked")
library(unmarked) #for occupancy analysis
#install.packages("AICcmodavg")
library(AICcmodavg) #for AICc models
#install.packages("data.table")
library(data.table) #For matrix formatting
#install.packages("ggalluvial")
library(ggalluvial) #For Sankey Diagram

setwd("~/Documents/GitHub/Garter_Snake_SR/GS_Analysis_Final")
```

### Datasets

```{r}
date = read.csv("Data_Tables/Dates.csv") #List of all surveyed dates

site = read.csv(file = "Data_Tables/Sites.csv") #List of all surveyed grids with lat/long

habitat <- read.csv(file = "Data_Tables/Hab_Covariates.csv") #habitat/site covariates at each grid

snake = read.csv(file = "Data_Tables/Snake_Capture_Data.csv") #records of individual captures of garter snakes

weather = read.csv(file = "Data_Tables/Newark_Airport_Weather.csv") #monthly weather data from the Licking County Regional Airport in Heath, Ohio from NOAA
```

## Exploratory Analysis and Data Formatting

These are diagnostic checks before running occupancy analysis. Covariates were collected in 2018, 2022 (fewer variables), 2019, and 2024.

### Site Covariate Correlation Plots

Checking to see if variables are too similar to prevent multicollinearity.

```{r}
#Separating habitat covariates by year
habitat_2024 <- habitat %>% 
  filter(Year == 2024) #filtering out 2024

habitat_2022 <- habitat %>% 
  filter(Year == 2022) #filtering out 2022

habitat_2019 <- habitat %>% 
  filter(Year == 2019) #filtering out 2019

habitat_2018 <- habitat %>% 
  filter(Year == 2018) #filtering out 2018

#Excluding categorical or missing columns - Pearson's correlation only runs with continuous variables. 
#Note: Distance variables should be binned due to some values being >50m, but for this, I will treat them as numerical variables to run correlation plots - there will be an underestimation leading to some bias, but this analysis is exploratory.
n_Habitat_2024 <- habitat_2024 %>% select(-Habitat_Type, -Grid, -Year)
n_Habitat_2022 <- habitat_2022 %>% select(Trees, Distance_to_Trail_m, Distance_to_Water_m)
n_Habitat_2019 <- habitat_2019 %>% select(-Habitat_Type, -Grid, -Year)
n_Habitat_2018 <- habitat_2018 %>% select(-Habitat_Type, -Grid, -Year)

#Pearson's correlation matrix for each year
hab_corr_2024 <- cor(n_Habitat_2024)
hab_corr_2022 <- cor(n_Habitat_2022)
hab_corr_2019 <- cor(n_Habitat_2019)
hab_corr_2018 <- cor(n_Habitat_2018)

#Correlation plots for habitat variables for each year
corrplot(hab_corr_2024, 
         type = "upper", 
         order = "original", 
         tl.col = "black", 
         tl.srt = 45, 
         main = "2024", 
         mar=c(1,1,1,1),
         family = "serif"
         )

corrplot(hab_corr_2022, 
         type = "upper", 
         order = "original", 
         tl.col = "black", 
         tl.srt = 45, 
         main = "2022", 
         mar=c(1,1,1,1),
         family = "serif"
         )

corrplot(hab_corr_2019, 
         type = "upper", 
         order = "original", 
         tl.col = "black", 
         tl.srt = 45, main = "2019", 
         mar=c(1,1,1,1),
         family = "serif"
         )

corrplot(hab_corr_2018, 
         type = "upper", 
         order = "original", 
         tl.col = "black", 
         tl.srt = 45, 
         main = "2018", 
         mar=c(1,1,1,1),
         family = "serif"
         )

```

::: callout-note
Since all the correlations are between 0.8 and -0.8, each variable can be run independently.
:::

### Site Covariate Boxplots

Checking to that fixed covariates (distance to water/distance to land) are the same across years and yearly covariates have variations throughout the years.

```{r}
#Boxplots for each covariate

ggplot(habitat, aes(x = Grid, y = Litter_Depth_cm)) + 
  geom_boxplot() +
  theme_minimal()
  
ggplot(habitat, aes(x = Grid, y = Max_Veg_Height_cm)) + 
  geom_boxplot() +
  theme_minimal()

ggplot(habitat, aes(x = Grid, y = Trees)) + 
  geom_boxplot() +
  theme_minimal()

ggplot(habitat, aes(x = Grid, y = Distance_to_Trail_m)) + 
  geom_boxplot() +
  theme_minimal() #same for all years

ggplot(habitat, aes(x = Grid, y = Distance_to_Water_m)) + 
  geom_boxplot() +
  theme_minimal() #same for all years

#The data spread for trees is really wide (especially since tree number should not change that much) mostly due to 2022 so making a new table without 2022
n_habitat <- habitat %>% 
  filter(Year %in% c(2018, 2019, 2024))

#Boxplot of number of trees without 2022
ggplot(n_habitat, aes(x = Grid, y = Trees)) + 
  geom_boxplot() +
  theme_minimal()
```

::: callout-note
Boxplots showed that some variables had a lot of variation, especially evident in trees, likely due to confusions between shrub and tree. Some of the original tree values were modified from original data in Grid G5 and E3 due to known miscounts of shrubs for trees, but there might be more inconsistencies in measurements across years that were not caught.

2022 especially had a high count of trees compared to other years and numbers across each of the 4 coverboards within a grid were the same rather than measured coverboard by coverboard, so 2022 will be omitted from using yearly variables in multi-season modelling.
:::

### Changes in Habitat Type over Time Plots

```{r}
#Tile plot of how Habitat Type changes for each coverboard across years
habitat %>%
  filter(Habitat_Type != "") %>% #since many years don't have measured hab type
  ggplot(aes(x = Year, y = Grid, fill = Habitat_Type)) +
  geom_tile(color = "black") + 
  scale_fill_manual( name = "Habitat Type",
                     values = c("Forb" = "#F2B50B", 
                               "Grass/Forb" = "#FFD8DF", 
                               "Grass" = "#D8E983", 
                               "Wetland" = "skyblue"),
                     ) +
  theme_minimal()

#Sankey plot of the number and trend in changes of Habitat Type across years
Hab_sankey_plot <- habitat %>%
  filter(Habitat_Type != "") %>% #since many years don't have measured hab type
  ggplot(aes(x = Year, stratum = Habitat_Type, alluvium = Grid, 
  fill = Habitat_Type)) +
  geom_flow(alpha = 0.6, na.rm = TRUE) +
  geom_stratum() + 
  scale_fill_manual( name = "Habitat Type",
                     values = c("Forb" = "#F2B50B", 
                               "Grass/Forb" = "#FFAAB8", 
                               "Grass" = "#D8E983", 
                               "Wetland" = "skyblue")) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18)
  )

Hab_sankey_plot

#saving image for thesis
ggsave("Hab_sankey_plot.png", plot = Hab_sankey_plot, width = 10, height = 6, units = "in", dpi = 800)
```

::: callout-note
Some of these changes over time might be due to inconsistencies in measurements with different observers.
:::

### Abundance Graph

```{r}

snake_n <- snake %>%
  filter(Year != 2016) #removing 2016

abundance_sum <- snake_n %>%
  group_by(Year) %>%
  summarise(Abundance = n(), .groups = "drop") #summing numbers for abundance by year

#Abundance plot
ggplot(abundance_sum, aes(x = Year, y = Abundance)) +
  geom_line(color = "mistyrose1", linewidth = 3) +
  geom_point(color = "mistyrose3", size = 4) + 
  scale_x_continuous(breaks = seq(min(abundance_sum$Year),
                                  max(abundance_sum$Year),
                                  by = 1)) + 
  scale_y_continuous(
    limits = c(0, NA)  #so y axis starts at 0
  ) +
  labs(
    x = "Year",
    y = "Abundance"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18)
  )
```

## Dataset Modification

Modifying datasets before analysis and making detection table.

### Binning a Distance Variable

```{r}
#Binning distance to water since there was already binned value (over 50m)
habitat <- habitat %>%
  mutate(Water_Bin = case_when(
    Distance_to_Water_m <= 10 ~ "0–10",
    Distance_to_Water_m <= 25 ~ "11–25",
    Distance_to_Water_m <= 49 ~ "26–49",
    Distance_to_Water_m == 50 ~ "50+"
  ))
#converting new binned variable to factor
habitat$Water_Bin <- as.factor(habitat$Water_Bin)

#checking the order is right
levels(habitat$Water_Bin)
```

### Table with Total Effort and Detections from Snake Capture Data

```{r}
#Making sure R reads the Full_Date column in both tables as a date
date$Full_Date <- as.Date(date$Full_Date, format = "%m/%d/%y")
snake$Full_Date <- as.Date(snake$Full_Date, format = "%m/%d/%y")

#Adding a column of month and year to bin into months for multi-season analysis
date <- date %>%
  mutate(Month_Year = format(Full_Date, "%m-%Y"))

#New table of total efforts of each checked date with all 29 grids using tidyr by adding every variable in date into effort_grid treating each row as a unit so nothing gets misaligned
effort_grid <- crossing(
  Grid = site$Grid,
  date
)

#Detection table using each row of effort_grid and snake capture data for 0/1 (presence-absence) by gridpoint by date
detection <- effort_grid %>%
  left_join(
    snake %>%
      distinct(Grid, Full_Date) %>%
      mutate(Detection = 1),
    by = c("Grid", "Full_Date")
  ) %>% 
  mutate(Detection = ifelse(is.na(Detection), 0, Detection))

head(detection)
```

::: callout-note
The 'detection' column is 0 if no snake was detected at the site and a 1 if at least 1 snake was detected at the site. All of the detection matrices required for occupancy analysis will be generated from this.
:::

## Single-Season Occupancy

Occupancy analysis with and without habitat covariates for each year. Every year will be run with the fixed distance covariates. 2022 will be run with number of trees within 10mas well. 2024, 2019, and 2018 will be run with all of the site covariates since only those years have full habitat surveys. The code will be repetitive.

### 2025

```{r}
#Filtering out only dates in 2025
detection25 <- detection %>% 
  filter(Year == 2025)

nrow(detection25) #should be 841 rows and there are 29 grids so 2025 has 29 sampling days

n_detection25 <- detection25 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix25 <- n_detection25 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2025
habitat_2025 <- habitat %>% 
  filter(Year == 2025)

#Null model
umf_25 <- unmarkedFrameOccu(
  y = det_matrix25,
  siteCovs = habitat_2025
) #converting to unmarked frame to use package 'unmarked'

summary(umf_25) #checking

#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_25 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_25)

summary(m0_25) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_25, type = "det")    #Detection
backTransform(m0_25, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_25, type = "det")) #Detection
confint(backTransform(m0_25, type = "state")) #Occupancy

#Detection only models
det_models_25 <- list(
  null = occu(~1 ~1, data = umf_25),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_25),
  d_water = occu(~Water_Bin ~1, data = umf_25),
  d_water_trail = occu(~Water_Bin+Distance_to_Trail_m ~1, data = umf_25)
  ) #List of models for 2025

aictab(cand.set = det_models_25) #Ranks models via aicc

#Adding fixed habitat covariates with null occupancy.
occu_models_25 <- list(
  o_null = occu(~1 ~1, data = umf_25),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_25),
  o_water = occu(~1 ~Water_Bin, data = umf_25),
  o_water_trail = occu(~1 ~Distance_to_Trail_m + Water_Bin, data = umf_25)
  ) #List of models for 2025

aictab(cand.set = occu_models_25) #Ranks models via aicc

```

### 2024

```{r}
#Filtering out only dates in 2024
detection24 <- detection %>% 
  filter(Year == 2024)

nrow(detection24) #should be 841 rows and there are 29 grids so 2024 has 29 sampling days

n_detection24 <- detection24 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix24 <- n_detection24 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2024
habitat_24 <- habitat %>% 
  filter(Year == 2024)

#Null model
umf_24 <- unmarkedFrameOccu(
  y = det_matrix24,
  siteCovs = habitat_24
) #converting to unmarked frame to use package 'unmarked'

summary(umf_24) #checking


#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_24 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_24)

summary(m0_24) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_24, type = "det")    #Detection
backTransform(m0_24, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_24, type = "det")) #Detection
confint(backTransform(m0_24, type = "state")) #Occupancy

#Testing fixed habitat covariates with detection
det_models_24 <- list(
  null = occu(~1 ~1, data = umf_24),
  d_veg = occu(~Max_Veg_Height_cm ~1, data = umf_24),
  d_tree = occu(~Trees ~1, data = umf_24),
  d_hab = occu(~Habitat_Type ~1, data = umf_24),
  d_litter = occu(~Litter_Depth_cm ~1, data = umf_24),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_24),
  d_water = occu(~Water_Bin ~1, data = umf_24)
)

aictab(cand.set = det_models_24) #Ranks models via aicc

#Testing fixed habitat covariates with occupancy.
occu_models_24 <- list(
  null = occu(~1 ~1, data = umf_24),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_24),
  o_water = occu(~1 ~Water_Bin, data = umf_24),
  o_habitat = occu(~1 ~Habitat_Type, data = umf_24),
  o_litter = occu(~1 ~Litter_Depth_cm, data = umf_24),
  o_veg = occu(~1 ~Max_Veg_Height_cm, data = umf_24),
  o_tree = occu(~1 ~Trees, data = umf_24)
  ) #List of models for 2024

aictab(cand.set = occu_models_24) #Ranks models via aicc

#Adding fixed habitat covariates with occupancy with best fit detection model: habitat type
models_24 <- list(
  null = occu(~1 ~1, data = umf_24),
  trail = occu(~Habitat_Type ~Distance_to_Trail_m, data = umf_24),
  water = occu(~Habitat_Type ~Water_Bin, data = umf_24),
  habitat = occu(~Habitat_Type ~Habitat_Type, data = umf_24),
  litter = occu(~Habitat_Type ~Litter_Depth_cm, data = umf_24),
  veg = occu(~Habitat_Type ~Max_Veg_Height_cm, data = umf_24),
  tree = occu(~Habitat_Type ~Trees, data = umf_24)
  ) #List of models for 2024

aictab(cand.set = models_24) #Ranks models via aicc

litter = occu(~Habitat_Type ~Litter_Depth_cm, data = umf_24)
litter
predict(litter, type = "state")
predict(litter, type = "det")
```

```{r}
#best fit model for 2024
best_24 = occu(~Habitat_Type ~Litter_Depth_cm, data = umf_24)

best_24 <- predict(best_24, type = "state")  #Occupancy

#Add grid as a column
best_24$Grid <- habitat_24$Grid
```

### 2023

```{r}
#Filtering out only dates in 2023
detection23 <- detection %>% 
  filter(Year == 2023)

nrow(detection23) #should be 812 rows and there are 29 grids so 2023 has 28 sampling days

n_detection23 <- detection23 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix23 <- n_detection23 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2023
habitat_23 <- habitat %>% 
  filter(Year == 2023)

#Null model
umf_23 <- unmarkedFrameOccu(
  y = det_matrix23,
  siteCovs = habitat_23
) #converting to unmarked frame to use package 'unmarked'

summary(umf_23) #checking

#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_23 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_23)

summary(m0_23) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_23, type = "det")    #Detection
backTransform(m0_23, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_23, type = "det")) #Detection
confint(backTransform(m0_23, type = "state")) #Occupancy

#Detection only models
det_models_23 <- list(
  null = occu(~1 ~1, data = umf_23),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_23),
  d_water = occu(~Water_Bin ~1, data = umf_23),
  d_water_trail = occu(~Water_Bin + Distance_to_Trail_m ~1, data = umf_23)
  ) #List of models for 2023

aictab(cand.set = det_models_23) #Ranks models via aicc

#Adding fixed habitat covariates with null occupancy.
occu_models_23 <- list(
  o_null = occu(~1 ~1, data = umf_23),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_23),
  o_water = occu(~1 ~Water_Bin, data = umf_23),
  o_water_trail = occu(~1 ~Distance_to_Trail_m + Water_Bin, data = umf_23)
  ) #List of models for 2023

aictab(cand.set = occu_models_23) #Ranks models via aicc
```

### 2022

```{r}
#Filtering out only dates in 2022
detection22 <- detection %>% 
  filter(Year == 2022)

nrow(detection22) #should be 754 rows and there are 29 grids so 2022 has 26 sampling days

n_detection22 <- detection22 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix22 <- n_detection22 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2022
habitat_22 <- habitat %>% 
  filter(Year == 2022)

#Null model
umf_22 <- unmarkedFrameOccu(
  y = det_matrix22,
  siteCovs = habitat_22
) #converting to unmarked frame to use package 'unmarked'

summary(umf_22) #checking

#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_22 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_22)

summary(m0_22) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_22, type = "det")    #Detection
backTransform(m0_22, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_22, type = "det")) #Detection
confint(backTransform(m0_22, type = "state")) #Occupancy

#Detection only models
det_models_22 <- list(
  null = occu(~1 ~1, data = umf_22),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_22),
  d_water = occu(~Water_Bin ~1, data = umf_22),
  d_water_trail_tree = occu(~Water_Bin+Distance_to_Trail_m+Trees ~1, data = umf_22),
  d_tree = occu(~Trees ~1, data = umf_22)
  ) #List of models for 2022

aictab(cand.set = det_models_22) #Ranks models via aicc

#Adding fixed habitat covariates with null occupancy.
occu_models_22 <- list(
  o_null = occu(~1 ~1, data = umf_22),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_22),
  o_water = occu(~1 ~Water_Bin, data = umf_22),
  o_tree = occu(~1 ~Trees, data = umf_22),
  o_water_trail_tree = occu(~1 ~Distance_to_Trail_m + Water_Bin + Trees, data = umf_22)
  ) #List of models for 2022

aictab(cand.set = occu_models_22) #Ranks models via aicc

#Adding fixed habitat covariates with best fit model: binned distance from water
models_22 <- list(
  null = occu(~1 ~1, data = umf_22),
  trail = occu(~Water_Bin ~Distance_to_Trail_m, data = umf_22),
  water = occu(~Water_Bin ~Water_Bin, data = umf_22),
  tree = occu(~Water_Bin ~Trees, data = umf_22),
  water_trail_tree = occu(~Water_Bin ~Distance_to_Trail_m + Water_Bin + Trees, data = umf_22)
  ) #List of models for 2022

aictab(cand.set = models_22) #Ranks models via aicc

tree = occu(~Water_Bin ~Trees, data = umf_22)
tree
```

```{r}
#best fit model for 2022
best_22 = occu(~Water_Bin ~Trees, data = umf_22)

best_22 <- predict(best_22, type = "state")  #Occupancy

#Add grid as a column
best_22$Grid <- habitat_22$Grid
```

### 2021

```{r}
#Filtering out only dates in 2021
detection21 <- detection %>% 
  filter(Year == 2021)

nrow(detection21) #should be 551 rows and there are 29 grids so 2021 has 19 sampling days

n_detection21 <- detection21 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix21 <- n_detection21 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2021
habitat_21 <- habitat %>% 
  filter(Year == 2021)

#Null model
umf_21 <- unmarkedFrameOccu(
  y = det_matrix21,
  siteCovs = habitat_21
) #converting to unmarked frame to use package 'unmarked'

summary(umf_21) #checking

#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_21 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_21)

summary(m0_21) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_21, type = "det")    #Detection
backTransform(m0_21, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_21, type = "det")) #Detection
confint(backTransform(m0_21, type = "state")) #Occupancy

#Detection only models
det_models_21 <- list(
  null = occu(~1 ~1, data = umf_21),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_21),
  d_water = occu(~Water_Bin ~1, data = umf_21),
  d_water_trail = occu(~Water_Bin+Distance_to_Trail_m ~1, data = umf_21)
  ) #List of models for 2021

aictab(cand.set = det_models_21) #Ranks models via aicc

#Adding fixed habitat covariates with null occupancy.
occu_models_21 <- list(
  o_null = occu(~1 ~1, data = umf_21),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_21),
  o_water = occu(~1 ~Water_Bin, data = umf_21),
  o_water_trail = occu(~1 ~Distance_to_Trail_m + Water_Bin, data = umf_21)
  ) #List of models for 2021

aictab(cand.set = occu_models_21) #Ranks models via aicc

#Adding fixed habitat covariates with best fit model: binned distance from water
models_21 <- list(
  null = occu(~1 ~1, data = umf_21),
  trail = occu(~Water_Bin ~Distance_to_Trail_m, data = umf_21),
  water = occu(~Water_Bin ~Water_Bin, data = umf_21),
  water_trail = occu(~Water_Bin ~Distance_to_Trail_m + Water_Bin, data = umf_21)
  ) #List of models for 2021

aictab(cand.set = models_21) #Ranks models via aicc

trail = occu(~Water_Bin ~Distance_to_Trail_m, data = umf_21)
trail
```

```{r}
#best fit model for 2021
best_21 = occu(~Water_Bin ~Distance_to_Trail_m, data = umf_21)

best_21 <- predict(best_21, type = "state")  #Occupancy

#Add grid as a column
best_21$Grid <- habitat_21$Grid
```

### 2020

```{r}
#Filtering out only dates in 2020
detection20 <- detection %>% 
  filter(Year == 2020)

nrow(detection20) #should be 899 rows and there are 29 grids so 2020 has 31 sampling days

n_detection20 <- detection20 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix20 <- n_detection20 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2020
habitat_20 <- habitat %>% 
  filter(Year == 2020)

#Null model
umf_20 <- unmarkedFrameOccu(
  y = det_matrix20,
  siteCovs = habitat_20
) #converting to unmarked frame to use package 'unmarked'

summary(umf_20) #checking

#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_20 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_20)

summary(m0_20) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_20, type = "det")    #Detection
backTransform(m0_20, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_20, type = "det")) #Detection
confint(backTransform(m0_20, type = "state")) #Occupancy

#Detection only models
det_models_20 <- list(
  null = occu(~1 ~1, data = umf_20),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_20),
  d_water = occu(~Water_Bin ~1, data = umf_20),
  d_water_trail = occu(~Water_Bin+Distance_to_Trail_m ~1, data = umf_20)
  ) #List of models for 2020

aictab(cand.set = det_models_20) #Ranks models via aicc

#Adding fixed habitat covariates with null occupancy.
occu_models_20 <- list(
  o_null = occu(~1 ~1, data = umf_20),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_20),
  o_water = occu(~1 ~Water_Bin, data = umf_20),
  o_water_trail = occu(~1 ~Distance_to_Trail_m + Water_Bin, data = umf_20)
  ) #List of models for 2020

aictab(cand.set = occu_models_20) #Ranks models via aicc

#Adding fixed habitat covariates with best fit model: binned distance from trail
models_20 <- list(
  null = occu(~1 ~1, data = umf_20),
  trail = occu(~Distance_to_Trail_m ~Distance_to_Trail_m, data = umf_20),
  water = occu(~Distance_to_Trail_m ~Water_Bin, data = umf_20),
  water_trail = occu(~Distance_to_Trail_m ~Distance_to_Trail_m + Water_Bin, data = umf_20)
  ) #List of models for 2020

aictab(cand.set = models_20) #Ranks models via aicc
```

### 2019

```{r}
#Filtering out only dates in 2019
detection19 <- detection %>% 
  filter(Year == 2019)

nrow(detection19) #should be 841 rows and there are 29 grids so 2019 has 29 sampling days

n_detection19 <- detection19 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix19 <- n_detection19 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2019
habitat_19 <- habitat %>% 
  filter(Year == 2019)

#Null model
umf_19 <- unmarkedFrameOccu(
  y = det_matrix19,
  siteCovs = habitat_19
) #converting to unmarked frame to use package 'unmarked'

summary(umf_19) #checking


#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_19 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_19)

summary(m0_19) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_19, type = "det")    #Detection
backTransform(m0_19, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_19, type = "det")) #Detection
confint(backTransform(m0_19, type = "state")) #Occupancy

#Testing fixed habitat covariates with detection
det_models_19 <- list(
  null = occu(~1 ~1, data = umf_19),
  d_veg = occu(~Max_Veg_Height_cm ~1, data = umf_19),
  d_tree = occu(~Trees ~1, data = umf_19),
  d_hab = occu(~Habitat_Type ~1, data = umf_19),
  d_litter = occu(~Litter_Depth_cm ~1, data = umf_19),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_19),
  d_water = occu(~Water_Bin ~1, data = umf_19)
)

aictab(cand.set = det_models_19) #Ranks models via aicc

#Testing fixed habitat covariates with occupancy.
occu_models_19 <- list(
  null = occu(~1 ~1, data = umf_19),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_19),
  o_water = occu(~1 ~Water_Bin, data = umf_19),
  o_habitat = occu(~1 ~Habitat_Type, data = umf_19),
  o_litter = occu(~1 ~Litter_Depth_cm, data = umf_19),
  o_veg = occu(~1 ~Max_Veg_Height_cm, data = umf_19),
  o_tree = occu(~1 ~Trees, data = umf_19)
  ) #List of models for 2019

aictab(cand.set = occu_models_19) #Ranks models via aicc

#Adding fixed habitat covariates with occupancy with best fit detection model: Distance to Trail
models_19 <- list(
  null = occu(~1 ~1, data = umf_19),
  trail = occu(~Distance_to_Trail_m ~Distance_to_Trail_m, data = umf_19),
  water = occu(~Distance_to_Trail_m ~Water_Bin, data = umf_19),
  habitat = occu(~Distance_to_Trail_m ~Habitat_Type, data = umf_19),
  litter = occu(~Distance_to_Trail_m ~Litter_Depth_cm, data = umf_19),
  veg = occu(~Distance_to_Trail_m ~Max_Veg_Height_cm, data = umf_19),
  tree = occu(~Distance_to_Trail_m ~Trees, data = umf_19)
  ) #List of models for 2019

aictab(cand.set = models_19) #Ranks models via aicc
```

### 2018

```{r}
#Filtering out only dates in 2018
detection18 <- detection %>% 
  filter(Year == 2018)

nrow(detection18) #should be 899 rows and there are 29 grids so 2018 has 31 sampling days

n_detection18 <- detection18 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix18 <- n_detection18 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2018
habitat_18 <- habitat %>% 
  filter(Year == 2018)

#Null model
umf_18 <- unmarkedFrameOccu(
  y = det_matrix18,
  siteCovs = habitat_18
) #converting to unmarked frame to use package 'unmarked'

summary(umf_18) #checking


#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_18 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_18)

summary(m0_18) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_18, type = "det")    #Detection
backTransform(m0_18, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_18, type = "det")) #Detection
confint(backTransform(m0_18, type = "state")) #Occupancy

#Testing fixed habitat covariates with detection
det_models_18 <- list(
  null = occu(~1 ~1, data = umf_18),
  d_veg = occu(~Max_Veg_Height_cm ~1, data = umf_18),
  d_tree = occu(~Trees ~1, data = umf_18),
  d_hab = occu(~Habitat_Type ~1, data = umf_18),
  d_litter = occu(~Litter_Depth_cm ~1, data = umf_18),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_18),
  d_water = occu(~Water_Bin ~1, data = umf_18)
)

aictab(cand.set = det_models_18) #Ranks models via aicc

#Testing fixed habitat covariates with occupancy.
occu_models_18 <- list(
  null = occu(~1 ~1, data = umf_18),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_18),
  o_water = occu(~1 ~Water_Bin, data = umf_18),
  o_habitat = occu(~1 ~Habitat_Type, data = umf_18),
  o_litter = occu(~1 ~Litter_Depth_cm, data = umf_18),
  o_veg = occu(~1 ~Max_Veg_Height_cm, data = umf_18),
  o_tree = occu(~1 ~Trees, data = umf_18)
  ) #List of models for 2018

aictab(cand.set = occu_models_18) #Ranks models via aicc

#Adding fixed habitat covariates with occupancy with second best fit detection model: Litter Depth
models_18 <- list(
  null = occu(~1 ~1, data = umf_18),
  trail = occu(~Litter_Depth_cm ~Distance_to_Trail_m, data = umf_18),
  water = occu(~Litter_Depth_cm ~Water_Bin, data = umf_18),
  habitat = occu(~Litter_Depth_cm ~Habitat_Type, data = umf_18),
  litter = occu(~Litter_Depth_cm ~Litter_Depth_cm, data = umf_18),
  veg = occu(~Litter_Depth_cm ~Max_Veg_Height_cm, data = umf_18),
  tree = occu(~Litter_Depth_cm ~Trees, data = umf_18)
  ) #List of models for 2018

aictab(cand.set = models_18) #Ranks models via aicc
```

### 2017

```{r}
#Filtering out only dates in 2017
detection17 <- detection %>% 
  filter(Year == 2017)

nrow(detection17) #should be 812 rows and there are 29 grids so 2017 has 28 sampling days

n_detection17 <- detection17 %>%
  select(Grid, Full_Date, Detection)

#Generates matrix from detection data where rows are grids and columns are dates
det_matrix17 <- n_detection17 %>%
  pivot_wider(
    names_from  = Full_Date,
    values_from = Detection
  ) %>%
  column_to_rownames("Grid") %>% #this is prefered structure for 'unmarked'
  as.matrix()

#Filtering out 2017
habitat_17 <- habitat %>% 
  filter(Year == 2017)

#Null model
umf_17 <- unmarkedFrameOccu(
  y = det_matrix17,
  siteCovs = habitat_17
) #converting to unmarked frame to use package 'unmarked'

summary(umf_17) #checking

#Running null occupancy model - assumes that detection and occupancy are the same across grids
m0_17 <- occu(formula = ~1 # detection formula first
                        ~1, # occupancy formula second, 
                  data = umf_17)

summary(m0_17) #Show AIC, and estimates, SE, z-scores (on logit scale)

#Changing from logit scale to probability using ψ=1/1+e^-(Logit value)
backTransform(m0_17, type = "det")    #Detection
backTransform(m0_17, type = "state")  #Occupancy

#Confidence intervals on probability scale
confint(backTransform(m0_17, type = "det")) #Detection
confint(backTransform(m0_17, type = "state")) #Occupancy

#Detection only models
det_models_17 <- list(
  null = occu(~1 ~1, data = umf_17),
  d_trail = occu(~Distance_to_Trail_m ~1, data = umf_17),
  d_water = occu(~Water_Bin ~1, data = umf_17),
  d_water_trail = occu(~Water_Bin+Distance_to_Trail_m ~1, data = umf_17)
  ) #List of models for 2017

aictab(cand.set = det_models_17) #Ranks models via aicc

#Adding fixed habitat covariates with null occupancy.
occu_models_17 <- list(
  o_null = occu(~1 ~1, data = umf_17),
  o_trail = occu(~1 ~Distance_to_Trail_m, data = umf_17),
  o_water = occu(~1 ~Water_Bin, data = umf_17),
  o_water_trail = occu(~1 ~Distance_to_Trail_m + Water_Bin, data = umf_17)
  ) #List of models for 2017

aictab(cand.set = occu_models_17) #Ranks models via aicc
```

### Naive Occupancy

```{r}
#Uses umf to find the exact proportion of sites with at least detection each year - naive occupancy
mean(rowSums(umf_25@y) > 0)
mean(rowSums(umf_24@y) > 0)
mean(rowSums(umf_23@y) > 0)
mean(rowSums(umf_22@y) > 0)
mean(rowSums(umf_21@y) > 0)
mean(rowSums(umf_20@y) > 0)
mean(rowSums(umf_19@y) > 0)
mean(rowSums(umf_18@y) > 0)
mean(rowSums(umf_17@y) > 0)
```

## Graphing Single Season Occupancy

### Null Occupancy Line Plot

```{r}
#Using this dataset
null_results <- read.csv("Data_Tables/Occupancy_Results.csv") #table of manually entered null occupancy and detection results from single-season analysis with naive ocupancy


#Line Plots
Plot_occu_line <- ggplot(null_results, aes(x = Year, y = Estimate, color = Parameter, fill = Parameter)) +
  geom_ribbon(data = null_results,
              aes(ymin = Lower, ymax = Upper),
              alpha = 0.2, color = NA, show.legend = FALSE) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 5, show.legend = FALSE) + 
  scale_color_manual(
  values = c("Detection" = "darkseagreen", "Naive_Occu" = "lightblue", "Pred_Occu" = "steelblue"),
  labels = c("Detection", "Naive Occupancy", "Predicted Occupancy")
) +
  scale_fill_manual(
    values = c("Pred_Occu" = "steelblue", "Detection" = "darkseagreen")
  ) +
  labs(x = "Year", y = "Probability", color = "Parameter", fill = "") +
  annotate(
  "text",
  x = min(null_results$Year),
  y = max(null_results$Estimate, na.rm = TRUE),
  label = "*",
  color = "red",
  size = 11,
  hjust = 0.5,
  vjust = 2.0
) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = sort(unique(null_results$Year))) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

ggsave("Plot_occu_Line.png", plot = Plot_occu_line, width = 10, height = 6, units = "in", dpi = 800)

Plot_occu_line
```

### Map of Gridpoint Occupancy with Best Fit Models

```{r}
library(sf)


#Defines lat/long boundaries for land_lab to use for map
land_lab <- st_bbox(c(
  xmin = -82.54700, #Long
  xmax = -82.54000,
  ymin = 40.08750, #Lat
  ymax = 40.09180
), crs = st_crs(4326))

#converting to 'simple feature' object
land_lab_sf <- st_as_sfc(land_lab)
land_lab_sf <- st_as_sf(land_lab_sf)

#importing a kmf file from manual Google Earth polygons/line drawings
ponds_paths <- st_read("Land Lab.kml")

#testing plot
plot(st_geometry(land_lab_sf), col = NA, border = "black") #land lab box
plot(st_geometry(ponds_paths), add = TRUE, col = "blue") #paths and ponds on top


#convert sites table to be readable by 'sf' library
site_sf <- st_as_sf(
  site,
  coords = c("Long", "Lat"),  #long 1st lat 2nd
  crs = 4326,
  remove = FALSE
)

#adding to land lab with gridpoints
site_map <- ggplot() +
  geom_sf(data = land_lab_sf, fill = NA, color = "black") +
  geom_sf(data = ponds_paths, fill = "blue", color = "darkblue") +
  geom_sf(data = site_sf, color = "red", size = 3) +
  coord_sf() +
  theme_minimal()

site_map

ponds_paths <- st_transform(ponds_paths, st_crs(land_lab_sf))
site_sf   <- st_transform(site_sf, st_crs(land_lab_sf))

#adding to graph
plot(st_geometry(land_lab_sf), col = NA, border = "black", main = "Land Lab with Ponds and Sites")
plot(st_geometry(ponds_paths), add = TRUE, col = "lightblue")
plot(st_geometry(site_sf), add = TRUE, col = "red", pch = 19)

#base plot with lat/long and labeled gridpoints
ggplot() +
  geom_sf(data = land_lab_sf, fill = NA, color = "black") +
  geom_sf(data = ponds_paths, fill = "lightblue", color = "darkblue") +
  geom_sf(data = site_sf, color = "red", size = 3) +
  geom_text(data = site, aes(x = Long, y = Lat, label = Grid), nudge_y = 0.0003) +
  coord_sf(xlim = c(land_lab["xmin"], land_lab["xmax"]), ylim = c(land_lab["ymin"], land_lab["ymax"])) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Map for best fit model in 2024

```{r}
#using this dataset containing predicted values for best model in 2024
best_24

#combining tables
grid_occu24 <- best_24 %>%
  left_join(site, by = "Grid")

#map
ggplot() +
  geom_sf(data = land_lab_sf, fill = NA, color = "black") +
  geom_sf(data = ponds_paths, fill = "lightblue", color = "darkblue") +
  # Points colored by Predicted occupancy
  geom_point(
    data = grid_occu24, 
    aes(x = Long, y = Lat, color = Predicted), 
    size = 5
  ) +
  geom_text(
    data = grid_occu24, 
    aes(x = Long, y = Lat, label = Grid), 
    nudge_y = 0.0003) +
  scale_color_gradient2(
    low = "blue", 
    mid = "white", 
    high = "red",  
    midpoint = 0.90) +
  coord_sf(
    xlim = c(land_lab["xmin"], land_lab["xmax"]), 
    ylim = c(land_lab["ymin"], land_lab["ymax"])
  ) +
  theme_minimal(base_family = "serif") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(color = "Predicted Occupancy")  # label for the gradient legend
```

### Map for best fit model in 2022

```{r}
#using this dataset containing predicted values for best model in 2024
best_22

#combining tables
grid_occu22 <- best_22 %>%
  left_join(site, by = "Grid")

#map
ggplot() +
  geom_sf(data = land_lab_sf, fill = NA, color = "black") +
  geom_sf(data = ponds_paths, fill = "lightblue", color = "darkblue") +
  # Points colored by Predicted occupancy
  geom_point(
    data = grid_occu22, 
    aes(x = Long, y = Lat, color = Predicted), 
    size = 5
  ) +
  geom_text(
    data = grid_occu22, 
    aes(x = Long, y = Lat, label = Grid), 
    nudge_y = 0.0003) +
  scale_color_gradient2(
    low = "blue", 
    mid = "white", 
    high = "red",  
    midpoint = 0.8) +
  coord_sf(
    xlim = c(land_lab["xmin"], land_lab["xmax"]), 
    ylim = c(land_lab["ymin"], land_lab["ymax"])
  ) +
  theme_minimal(base_family = "serif") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(color = "Predicted Occupancy")  # label for the gradient legend
```

### Map for best fit model in 2021

```{r}
#using this dataset containing predicted values for best model in 2021
best_21

#combining tables
grid_occu21 <- best_21 %>%
  left_join(site, by = "Grid")

#map
ggplot() +
  geom_sf(data = land_lab_sf, fill = NA, color = "black") +
  geom_sf(data = ponds_paths, fill = "lightblue", color = "darkblue") +
  # Points colored by Predicted occupancy
  geom_point(
    data = grid_occu21, 
    aes(x = Long, y = Lat, color = Predicted), 
    size = 5
  ) +
  geom_text(
    data = grid_occu21, 
    aes(x = Long, y = Lat, label = Grid), 
    nudge_y = 0.0003) +
  scale_color_gradient2(
    low = "blue", 
    mid = "white", 
    high = "red",  
    midpoint = 0.5) +
  coord_sf(
    xlim = c(land_lab["xmin"], land_lab["xmax"]), 
    ylim = c(land_lab["ymin"], land_lab["ymax"])
  ) +
  theme_minimal(base_family = "serif") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(color = "Predicted Occupancy")  # label for the gradient legend
```

## Multi-Season Occupancy

Dynamic occupancy analysis from 2017-2025.

Includes colonization probability (𝛾𝑖), extinction probability (𝜖𝑖) in addition to detection probability (𝑝𝑖) and initial occupancy (Ψ𝑖),

```{r}

m_date <- date %>% 
  subset(select = c(Year, Month_Year, Month)) %>% 
  distinct(Month_Year, .keep_all = TRUE) #keeping only distinct month_years without dates

hab_cov <- habitat %>% 
  subset(select = c(Grid, Distance_to_Trail_m, Water_Bin)) %>% 
  distinct(Grid, .keep_all = TRUE) #subsets only the site covariates that are consistent across years
yearly_hab_cov <- habitat %>% 
  subset(select = -c(Distance_to_Trail_m, Water_Bin)) #subsets only the site covariates that change from year to year

m_weather <- weather %>%
  mutate(Month = substr(Date, 6, 7), Year  = substr(Date, 1, 4), Month_Year = paste(Month, Year, sep = "-")
  ) #Turning date into Month-Year format

m_weather$Month <- month(mdy(m_weather$Month_Year))

m_weather <- m_weather %>%
  filter(as.integer(substr(Date, 6, 7)) %in% 4:10) #keeping only weather data from april to october

#Taking out 2016 and months outside of 4-10
detection <- detection %>%
  filter(Year != 2016) %>%
  filter(Month >= 4, Month <= 10)

#condensing all month-days by gridpoint
m_detection <- detection  %>%
  group_by(Grid, Month_Year) %>%
  summarise(Detection = max(Detection), .groups = "drop") #choose highest number to keep (so keep 1 for any months with at least one day labeled 1)
```

### Matrix Generation

Generating lots of matrices since all inputs are matrices instead of table columns.

Detection

```{r}
#combining the site cov and detection
m_detection_hab <- left_join(m_detection, hab_cov, by = "Grid")

#Turning into detection matrix
m_det_matrix <- m_detection_hab %>%
  pivot_wider(
    names_from  = Month_Year,
    values_from = Detection
  ) %>%
  as.matrix()

dim(m_det_matrix) #checking column numbers

#For det matrix - ordering the detection date columns chronologically - using lubridate
ordered_dates <- colnames(m_det_matrix[, 4:66]) #getting only the detection columns 
ordered_dates <- order(ordered_dates) #ordering the dates
ord_det_matrix <- m_det_matrix[, c(1:3, 3 + ordered_dates)]  #keep first 3 columns and reorder
```

Primary & Secondary Periods

```{r}
#sampling month and year matrix

m_month_grid <- crossing(
  Grid = site$Grid, m_date) #assigning all the grids to each month

m_month_grid %>%
  count(Grid) #sanity check


setDT(m_month_grid) #making into a usable format for library data.table

m_month_matrix <- m_month_grid %>% 
  dcast(Grid ~ Month_Year,
        value.var = "Month") #matrix-like table of grid by date filled with months
m_month_matrix <- as.matrix(m_month_matrix) #changing into proper matrix to reorder
ord_month_matrix <- m_month_matrix[, c(1, 1 + ordered_dates)]  #keep first column and reordering the dates for month matrix 

colnames(ord_month_matrix)[-1] <- paste0("secondary", seq_along(colnames(ord_month_matrix)[-1])) #changing column names after grid to "secondary#"

m_year_matrix <- m_month_grid %>% 
  dcast(Grid ~ Month_Year,
        value.var = "Year") #matrix-like table of grid by date filled with years
m_year_matrix <- as.matrix(m_year_matrix) #changing into proper matrix to reorder
ord_year_matrix <- m_year_matrix[, c(1, 1 + ordered_dates)]  #keep first 3 columns and reordering the dates for year matrix 
colnames(ord_year_matrix)[-1] <- paste0("primary", seq_along(colnames(ord_year_matrix)[-1])) #changing column names after grid to "primary#"

#converting back to df to use leftjoin
m_det_df <- as.data.frame(ord_det_matrix)
m_month_df <- as.data.frame(ord_month_matrix)
m_year_df <- as.data.frame(ord_year_matrix)

#combining all 3 (detection with site cov + month + year)
m_full_df <- m_det_df %>% 
  left_join(m_month_df, by = "Grid") %>% 
  left_join(m_year_df, by = "Grid")

dim(m_full_df) #sanity check: should have 29 columns (grid) and 66+63+63= 192 rows
##This is so I can have fewer datasets that can be started with in the final version of code instead of doing all the dataset manipulation before this
```

### Separating out Detection & Periods

A lot of following code is based off https://darinjmcneil.weebly.com/multi-season-occupancy.html and https://jamesepaterson.github.io/jamespatersonblog/2021-01-01_dynamicoccupancy.html). Secondary period included in observation covariate later.

```{r}
#checking rows using index to know where to group
index_full_df <- data.frame(Index = 1:ncol(m_full_df), Name = colnames(m_full_df))

rownames(m_full_df) <- m_full_df$Grid #making grid the rows

detection_matrix <- as.matrix(m_full_df[, 4:66]) #matrix of detection

sec_matrix <- as.matrix(m_full_df[, 67:129]) #matrix of secondary period

prim_matrix <- as.matrix(m_full_df[, 130:192]) #matrix of primary period

#sanity checks: all should be the same
dim(detection_matrix) 
dim(sec_matrix)
dim(prim_matrix)
```

### fixed siteCovs

Habitat covariates that are consistent across years

```{r}
#inserting directly into unmarked frame instead of making matrices because it was causing issues

#trail_matrix <- as.matrix(m_full_df[, 2, drop = FALSE]) #not using

#water_matrix <- as.matrix(m_full_df[, 3, drop = FALSE]) #not using
  

```

::: callout-note
Not including yearly site covariates in multi-season analysis since only 3 years were measured, and I ran into some errors. 3 primary periods (years) with 7 secondary periods in each (months) might not be enough.
:::

### obsCovs matrices

Additional observation covariates that could change by survey by site. In this case, will remain consistent for all the coverboards in each month and the weather data from Newark airport is nearest airport with enough consistent historical data.

```{r}
#prepping weather data

n_weather <- m_weather %>% 
  select(Month_Year, PRCP, TAVG, TMIN, TMAX, DT32, DX70) #selecting weather variables to use

n_weather <- n_weather %>% 
  left_join(m_month_grid, by = "Month_Year") #adding grid to weather data

#average precipitation matrix by secondary period
avgprecip_matrix <- n_weather %>% 
  pivot_wider(
    id_cols = Grid,          
    names_from = Month_Year,         
    values_from = PRCP) %>% 
  column_to_rownames("Grid") %>% 
  as.matrix()

#average temp matrix by secondary period
avgtemp_matrix <- n_weather %>% 
  pivot_wider(
    id_cols = Grid,          
    names_from = Month_Year,         
    values_from = TAVG) %>% 
  column_to_rownames("Grid") %>% 
  as.matrix()

#min temp matrix by secondary period
mintemp_matrix <- n_weather %>% 
  pivot_wider(
    id_cols = Grid,          
    names_from = Month_Year,         
    values_from = TMIN) %>% 
  column_to_rownames("Grid") %>% 
  as.matrix()

#max temp matrix by secondary period
maxtemp_matrix <- n_weather %>% 
  pivot_wider(
    id_cols = Grid,          
    names_from = Month_Year,         
    values_from = TMAX) %>% 
  column_to_rownames("Grid") %>% 
  as.matrix()

#min temp under 0 C matrix by secondary period
mintemp0_matrix <- n_weather %>% 
  pivot_wider(
    id_cols = Grid,          
    names_from = Month_Year,         
    values_from = DT32) %>% 
  column_to_rownames("Grid") %>% 
  as.matrix()

#max temp over 21.1 C matrix by secondary period
maxtemp21.1_matrix <- n_weather %>% 
  pivot_wider(
    id_cols = Grid,          
    names_from = Month_Year,         
    values_from = DX70) %>% 
  column_to_rownames("Grid") %>% 
  as.matrix()

#sanity checks - should be all the same
dim(avgprecip_matrix)
dim(avgtemp_matrix)
dim(mintemp_matrix)
dim(maxtemp_matrix)
dim(mintemp0_matrix)
dim(maxtemp21.1_matrix)
```

::: callout-note
List of all matrix:

detection_matrix (presences/absences)

sec_matrix, prim_matrix (secondary and primary period)

trail_matrix, water_matrix (siteCovs)

avgprecip_matrix, mintemp_matrix, maxtemp_matrix，mintemp0_matrix, maxtemp21.1_matrix (obsCovs)
:::

```{r}
#making unmarkedframe for occupancy analysis (including all matrices from above)
umf_multi <- unmarkedMultFrame(
  y = detection_matrix,
  obsCovs = list(
    sec = sec_matrix,
    avgPrecip = avgprecip_matrix,
    minTemp = mintemp_matrix,
    maxTemp = maxtemp_matrix,
    minTemp0 = as.matrix(mintemp0_matrix),
    maxTemp21.1 = as.matrix(maxtemp21.1_matrix)),
  numPrimary = 9) #9 is number of years

siteCovs(umf_multi) <- data.frame(
  trail = as.numeric(m_full_df[, 2]),
  water = as.factor(m_full_df[, 3])
)


#checking
str(siteCovs(umf_multi))
str(obsCovs(umf_multi))
```

### Running Multi-Season

```{r}
# knitr::opts_chunk$set(error = TRUE)
# #Building detection models
# 
# p.null <- colext(psiformula = ~1, #initial occupancy
#                  gammaformula = ~ 1, #colonization
#                  epsilonformula = ~ 1, #extinction
#                  pformula = ~ 1, #detection
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.sec <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ sec,
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.avgPrecip <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ avgPrecip,
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.minTemp <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ minTemp,
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.maxTemp <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ maxTemp,
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.min0 <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ minTemp0,
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.max21.1 <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ maxTemp21.1,
#                  data = umf_multi,
#                  method="BFGS")
# 
# p.full <- colext(psiformula= ~ 1,
#                  gammaformula = ~ 1,
#                  epsilonformula = ~ 1,
#                  pformula = ~ avgPrecip + minTemp + maxTemp + minTemp0 + maxTemp21.1,
#                  data = umf_multi,
#                  method="BFGS")
# 
# #ranking by AICc
# multi_list1 <- list(p.trail=p.trail,
#                p.water=p.water,
#                p.avgPrecip=p.avgPrecip,
#                p.minTemp=p.minTemp,
#                p.maxTemp=p.maxTemp,
#                p.min0=p.min0,
#                p.max21.1=p.max21.1,
#                p.null=p.null,
#                p.full=p.full
#                )
# aictab(multi_list1)
# 
# str(obsCovs(umf1))

```

These ran without an error but it prevented the html file from rendering. Possibly something on the back end.

```{r}
# knitr::opts_chunk$set(error = TRUE)
# #adding best fit model (mintemp)
# #Building models
# 
# mm_null <- colext(psiformula = ~1, #initial occupancy
#                  gammaformula = ~ 1, #colonization
#                  epsilonformula = ~ 1, #extinction
#                  pformula = ~ 1, #detection
#                  data = umf1, 
#                  method="BFGS")
# 
# mm_trail <- colext(psiformula= ~ trail, 
#                  gammaformula = ~ 1, 
#                  epsilonformula = ~ 1, 
#                  pformula = ~ minTemp, 
#                  data = umf1, 
#                  method="BFGS")
# 
# mm_water <- colext(psiformula= ~ water, 
#                 gammaformula = ~ 1, 
#                 epsilonformula = ~ 1, 
#                 pformula = ~ minTemp, 
#                 data = umf1, 
#                 method="BFGS")
# 
# mm_trail_water <- colext(psiformula= ~ trail+water, 
#                 gammaformula = ~ 1, 
#                 epsilonformula = ~ 1, 
#                 pformula = ~ minTemp, 
#                 data = umf1, 
#                 method="BFGS")
# 
# multi_list2 <- list(mm_null=mm_null,
#                mm_trail=mm_trail,
#                mm_water=mm_water,
#                mm_trail_water=mm_trail_water
#                )
# aictab(multi_list2)
#
#model with highest aicc ranking
#backTransform(mm_trail, type = "col") #colonization
#backTransform(mm_trail, type = "ext") #extinction

write.csv(best_21, "Qiu_GS2021_Occupancy_By_Grid.csv", row.names = FALSE)
write.csv(best_22, "Qiu_GS2022_Occupancy_By_Grid.csv", row.names = FALSE)
write.csv(best_24, "Qiu_GS2024_Occupancy_By_Grid.csv", row.names = FALSE)
```

::: callout-note
I decided not to use much of the multi-season results because it gave a lot of NaN as well so it would be difficult to say anything meaningful. The html not rendering for last 2 chunks might mean that something is going wrong in the back end, so results might be inaccurate.

Results for backtransforms of mm_trail model:

Backtransformed linear combination(s) of Colonization estimate(s) Estimate SE LinComb (Intercept) 0.465 0.0703 -0.142 1

Transformation: logistic Backtransformed linear combination(s) of Extinction estimate(s)

Estimate SE LinComb (Intercept) 0.0725 0.0281 -2.55 1
:::
